Longitudinal models

### 2024-11-15 5_longitudinal.qmd
paper

---
title: "Longitudinal models"
format: html
---

Questions: 
(from Nilam)

1. Is the mean of the latent “language ability” factor scores different across 1st assessment, 2nd assessment, 3rd assessment, etc? This is basically an ANOVA on the latent factor scores – using “assessment number” as the categorical predictor. – an examination of “practice effects”.

Is the mean of the latent “language ability” factor scores increasing with age?
- This is basically a regression on the latent factor scores – and a growth model.
- In which case we also implicitly test research questions about the variances
- Is the (between-person) variance of the latent “language ability” factor scores increasing with age?

- A more nuanced version of the “practice effects" examination might use delta_t. 
- Is the mean of the latent “language ability” factor different across retest interval?
- This is basically a regression on the latent factor scores.

- In which case we also implicitly test research questions about the variances
- Is the (between-person) variance of the latent “language ability” factor scores increasing with age?

3. If we are interested in the covariances –
- Is the structure of the latent “language ability” factor changing with age?
- This again gets us into a multi-group SEM –
- Testing whether the factor loading structure differs across groups.
- This is most easily invoked with using categorical “time variable” – e.g, assessment number, age-group.
- However, it is also possible to do using a continuous time variable (e.g., age). In this case we basically use a moving window to obtain the covariances and then basically use age as a moderator of the factor loadings to examine the age x factor loading interaction.

# Descriptives

## Longitudinal data

@@

ggplot(d_sub_long, 
       aes(x = delta_t, fill = as.factor(admin_num))) +
  geom_histogram() + 
  facet_wrap(~dataset_name)
```
## Datasets with CDIs

Longitudinal data with CDI is just FMW2013 right now.

```{r}
fmw <- filter(d_sub_long, dataset_name == "fmw_2013") |>
  mutate(t = ifelse(age < 23, "t1","t2")) |>
  group_by(subject_id, t) |>
  summarise(age = mean(age), 
            log_rt = mean(log_rt), 
            prod = mean(prod), 
            long_window_accuracy = mean(long_window_accuracy),
            delta_t = mean(delta_t)) |>
  pivot_wider(names_from = "t", 
              values_from = c("age","log_rt","prod","long_window_accuracy", "delta_t"))

summary(lm(prod_t2 ~ long_window_accuracy_t1 * log_rt_t1 + age_t2, 
           data = fmw))
```

## Without CDIs

```{r}
ggplot(d_sub_long, 
       aes(x = age, y = long_window_accuracy, group = subject_id, col = dataset_name)) +
  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
  geom_smooth(aes(group = 1)) + 
  scale_y_log10() + 
  geom_hline(lty = 2, yintercept = .5)

@@

```{r}
ggplot(filter(d_sub_long, n_trials_rt > 3),
       aes(x = age, y = log_rt, group = subject_id, col = dataset_name)) +
  geom_line(alpha = .3, position = position_jitter(width = .1, height = 0)) + 
  geom_smooth(aes(group = 1)) + 
  scale_y_log10() 
```
# Q1: age-related change in the latent score

Growth in latent language factor. This is the same as the previous factor analysis, but we only look within child and within dataset.

@@

```{r}
ggplot(d_sub |>
         filter(subject_id %in% longitudinal$subject_id), 
       aes(x = age, y = f1, group = subject_id, col = dataset_name)) + 
  geom_point(alpha = .4) + 
  geom_line(alpha = .4) + 
  geom_smooth(aes(group = dataset_name), method = "lm")
```
Note that nesting age slopes doesn't converge. Could go Bayesian if we cared...

@@

# Q2: relations between latents and other factors

Longitudinal model.

```{r}
lmod <- lmer(accuracy ~ age + speed_0 * delta_t + (1 | subject_id) + (1 | dataset_name), 
     data = d_long)

summary(lmod)
```

```{r}
ggplot(d_long, 
       aes(x = age, y = accuracy, group = subject_id, col = dataset_name, lty = speed_0_high)) + 
  geom_line(alpha = .4) + 
  geom_smooth(aes(group = speed_0_high), method = "lm") + 
  facet_wrap(~dataset_name, scales="free_x")
```

>>>>

---
title: "Longitudinal models"
format: 
  html:
    toc: true
execute: 
  cache: true
---

Notes from Nilam:

- don't z score, mean center + scale by the variance in one age group (e.g., youngest or oldest)
- check data on Draft 2: latent change scores
- propagate changes to D6 top model to bottom
- consider adding shared parameter to ensure invariance across factors at t1 and t2 for D6 model 1
- what is residual variance on each of the time points - if you say they are equal then it's measurement error, but if you let them vary, it's deviation
- fit with shared error across all of the observed -->>
Look at this part of the latent growth tutorial #observed residual variances (constrained to equality)- math2 ~~ sigma2_u*math2

the two-wave model:
change in inter-individual differences
vs. differences in intra-individual change

This markdown documents attempts to do longitudinal models for the Peekbank data.

```{r}
library(here)
source(here("helper","common.R"))
library(lavaan)
library(tidySEM)
d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))

```{r}
d_sub_mat <- d_sub |>
  ungroup() |>
  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age)

d_sub_mat_s <- d_sub_mat |>
  ungroup() |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
                         "long_window_acc_var", "prod", "comp")), 
                       ~ age_scale(.x, age))) 
```

# Descriptives of the longitudinal data

@@

ggplot(d_sub_long, 
       aes(x = delta_t, fill = as.factor(admin_num))) +
  geom_histogram(binwidth = 1) + 
  facet_wrap(~dataset_name)

@@

How much do we have production vs. comprehension vocab?

```{r}
d_long |>
  group_by(dataset_name, subject_id) |>
  summarise(n_prod = sum(!is.na(prod), na.rm=TRUE), 
            n_comp = sum(!is.na(comp), na.rm=TRUE)) |>
  summarise(n_prod = mean(n_prod, na.rm=TRUE), 
            n_comp = mean(n_comp, na.rm=TRUE))
```
# Q0: longitudinal stability of measures

```{r}
MONTH_CUTOFF <- 3

d_reliability <- d_long |>
  filter(delta_t <= MONTH_CUTOFF) |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "admin_num",
              values_from = c("rt", "rt_var", "long_window_accuracy", 
                              "long_window_acc_var", "prod", "comp"))

d_reliability |>
  summarise(rt_corr = cor(rt_1, rt_2, use = "pairwise.complete.obs"),
            rt_var_corr = cor(rt_var_1, rt_var_2, use = "pairwise.complete.obs"),
            acc_corr = cor(long_window_accuracy_1, 
                          long_window_accuracy_2, use = "pairwise.complete.obs"),
            acc_var_corr = cor(long_window_acc_var_1, 
                              long_window_acc_var_2, use = "pairwise.complete.obs"),
            prod_corr = cor(prod_1, prod_2, use = "pairwise.complete.obs"),
            comp_corr = cor(comp_1, comp_2, use = "pairwise.complete.obs"))
```

# Q1: age-related changes in latent scores

Let's look at growth in latent language factor.

This is the same as the previous factor analysis, but we only look within child and within dataset.

Summary - this is very interpretable and shows us that we see linear growth in the latents with the log of age.

Downside is that we don't get a sense of the relations between the latents over time.

@@

```{r}
ggplot(d_sub |>
         filter(subject_id %in% longitudinal$subject_id), 
       aes(x = age, y = f1, group = subject_id)) + 
  geom_point(alpha = .2) + 
  geom_line(alpha = .2) + 
  geom_smooth(aes(group = dataset_name), method = "lm") + 
  facet_wrap(~dataset_name)
```
Note that nesting age slopes doesn't converge. Could go Bayesian if we cared...

@@

## 3 factor approach

Let's try looking at the speed/accuracy/vocab model across time.

```{r}
fa3_model <- "vocab =~ prod + comp
              accuracy =~ long_window_accuracy + long_window_acc_var
              speed =~ rt + rt_var"

fit3 <- cfa(fa3_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

preds <- predict(fit3)

d_sub$vocab <- preds[,1]
d_sub$accuracy <- preds[,2]
d_sub$speed <- preds[,3]
```

```{r}
d_long <- d_sub |>
  filter(subject_id %in% longitudinal$subject_id) |>
  group_by(subject_id) |>
  arrange(administration_id) |>
  mutate(admin_num = 1:n(), 
         delta_t = c(0,diff(age)), 
         speed_0 = speed[1]) |>
  group_by(dataset_name) |>
  mutate(speed_0_slow = speed_0 > median(speed_0, na.rm=TRUE))
```

```{r}
ggplot(d_long |>
         pivot_longer(vocab:accuracy, names_to = "variable", values_to = "value"), 
       aes(x = age, y = value, group = subject_id, col = dataset_name)) + 
  # geom_point(alpha = .1) + 
  geom_line(alpha = .1) + 
  geom_smooth(aes(group = dataset_name), method = "lm") + 
  scale_x_log10() +
  facet_wrap(~variable) +
  theme(legend.position = "bottom") + 
  xlab("Age (months)") + 
  ylab("Latent score") 
```
Change in the latents over age.

```{r}
long_mod_acc <- lmer(accuracy ~ age + (1 | subject_id) + (1 | dataset_name), 
                     data = d_sub)
long_mod_speed <- lmer(speed ~ age + (1 | subject_id) + (1 | dataset_name), 
                       data = d_sub)
long_mod_vocab <- lmer(vocab ~ age + (1 | subject_id) + (1 | dataset_name), 
                       data = d_sub)

summary(long_mod_acc)
summary(long_mod_speed)
summary(long_mod_vocab)
```

# Q2: Does RT relate to growth in vocabulary?

We see from the factor analysis that we are getting a coherent speed factor, but does it relate to anything else over time?

## Draft 1: with LMERs

Longitudinal model of change in accuracy. This is pretty crude, I just wanted to try it.

```{r}
lmod_acc <- lmer(accuracy ~ age + speed_0 * delta_t + (1 | subject_id) + (1 | dataset_name), 
                 data = d_long)

summary(lmod_acc)
```

So speed at time zero does have an effect on accuracy. Low speed at t0 means accuracy is higher. And this interacts with delta_t, but isn't the coefficient in the wrong direction?

Here's a visualization that doesn't quite capture this.

```{r}
ggplot(filter(d_long, !is.na(speed_0_slow)),
       aes(x = age, y = accuracy, group = subject_id, 
           col = dataset_name, 
           lty = speed_0_slow)) + 
  geom_line(alpha = .2) + 
  geom_smooth(aes(group = speed_0_slow), method = "lm") + 
  facet_wrap(~dataset_name, scales="free_x")
```

Longitudinal model of change in vocabulary now.

```{r}
lmod_vocab <- lmer(vocab ~ age + speed_0 * delta_t + (1 | subject_id) + (1 | dataset_name), 
                   data = d_long)

summary(lmod_vocab)
```
We don't see differences in vocab over time (no interaction of delta_t).

```{r}
ggplot(d_long, 
       aes(x = age, y = vocab, group = subject_id, 
           col = dataset_name, 
           lty = speed_0_slow)) + 
  geom_line(alpha = .2) + 
  geom_smooth(aes(group = speed_0_slow), method = "lm") + 
  facet_wrap(~dataset_name, scales="free_x")
```

## Draft 2: latent change scores

Idea behind this approach is to get a delta-vocabulary score for the longitudinal data and look at its relationship to vocab.

```{r}
d_long_s <- d_long |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
                         "long_window_acc_var", "prod", "comp")), 
                ~ age_scale(.x, age))) 
```

First fit the three-factor model to just the first obs of the longitudinal data.

```{r}
fa3l_model <- "vocab_l =~ prod + comp
              accuracy_l =~ long_window_accuracy + long_window_acc_var
              speed_l =~ rt + rt_var"

fit3l <- cfa(fa3l_model, 
             data = filter(d_long_s, admin_num == 1), 
             std.lv=TRUE, missing='fiml')

summary(fit3l, fit.measures=TRUE, standardize=TRUE)
```
```{r}
layout_l = matrix(nrow=2, ncol = 6,
                  data = c("speed_l",NA,"accuracy_l",NA,"vocab_l",NA,
                           "rt_var","rt", "long_window_accuracy",
                           "long_window_acc_var", "prod","comp"), 
                  byrow = TRUE)

graph_sem(model = fit3l, text_size = 2, layout = layout_l)

Now let's compare to the three-factor model again, with the fit to delta-prod and delta-comp.

```{r}
slopes <- d_long |>
  group_by(dataset_name, subject_id) |>
  nest() |>
  mutate(delta_prod = map_dbl(data, 
                              possibly(\(d) coef(lm(prod ~ delta_t, data = d))[2], 
                                       otherwise = NA)), 
         delta_comp = map_dbl(data, 
                              possibly(\(d) coef(lm(comp ~ delta_t, data = d))[2], 
                                       otherwise = NA))) |>
  select(-data)

d_long_s <- left_join(d_long_s, slopes)

fa3l_delta_model <- "vocab_l =~ delta_prod + delta_comp
              accuracy_l =~ long_window_accuracy + long_window_acc_var
              speed_l =~ rt + rt_var"

fit3l_delta <- cfa(fa3l_delta_model, 
                   data = filter(d_long_s, admin_num == 1), 
                   std.lv=TRUE, missing='fiml')

summary(fit3l_delta, fit.measures=TRUE, standardize=TRUE)
```

```{r}
layout_ld = matrix(nrow=2, ncol = 6,
                   data = c("speed_l",NA,"accuracy_l",NA,"vocab_l",NA,
                            "rt_var","rt", "long_window_accuracy",
                            "long_window_acc_var", "delta_prod","delta_comp"), 
                   byrow = TRUE)
graph_sem(model = fit3l_delta, text_size = 2, layout = layout_ld)
```

## Draft 3: discrete variables

Need to create discrete variables for each thing changing across time.

Nice piece is that we can do this with all the data.

```{r}
d_sub_wide_mod_s <- d_sub |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
                         "long_window_acc_var", "prod", "comp")),
                ~ age_scale(.x, age)))

wide_rts <- d_sub_wide_mod_s |>
  select(dataset_name, subject_id, age, rt) |>
  mutate(age = floor(age)) |>
  pivot_wider(names_from = "age", values_from = "rt", 
              names_prefix = "rt_", 
              values_fn = mean)

wide_accs <- d_sub_wide_mod_s |>
  select(dataset_name, subject_id, age, long_window_accuracy) |>
  mutate(age = floor(age)) |>
  pivot_wider(names_from = "age", values_from = "long_window_accuracy", 
              names_prefix = "acc_",
              values_fn = mean)

wide_prods <- d_sub_wide_mod_s |>
  select(dataset_name, subject_id, age, prod) |>
  mutate(age = floor(age)) |>
  pivot_wider(names_from = "age", values_from = "prod", 
              names_prefix = "prod_",
              values_fn = mean)

wide_comps <- d_sub_wide_mod_s |>
  select(dataset_name, subject_id, age, comp) |>
  mutate(age = floor(age)) |>
  pivot_wider(names_from = "age", values_from = "comp", 
              names_prefix = "comp_",
              values_fn = mean)

d_wide <- full_join(wide_rts, wide_accs) |>
  full_join(wide_prods) |>
  full_join(wide_comps)
```

```{r}
ggplot(filter(d_sub_wide_mod_s, subject_id %in% longitudinal$subject_id),
       aes(x = rt, y = long_window_accuracy, group = subject_id)) + 
  geom_point(alpha = .2, aes(col = as.factor(subject_id)))+ 
  geom_line(alpha = .2, aes(col = as.factor(subject_id))) +
  geom_smooth(aes(group = 1), method = "lm") + 
  scale_color_discrete(guide = FALSE)
```

```{r}
coupled_change <- "accuracy_intercept =~ 1 * acc_13 
accuracy_intercept =~ 1 * acc_14 
accuracy_intercept =~ 1 * acc_15 
accuracy_intercept =~ 1 * acc_16 
accuracy_intercept =~ 1 * acc_17
accuracy_intercept =~ 1 * acc_18 
accuracy_intercept =~ 1 * acc_19
accuracy_intercept =~ 1 * acc_20
accuracy_intercept =~ 1 * acc_21
accuracy_intercept =~ 1 * acc_22
accuracy_intercept =~ 1 * acc_23
accuracy_intercept =~ 1 * acc_24
accuracy_intercept =~ 1 * acc_25
accuracy_intercept =~ 1 * acc_26

accuracy_slope =~ 0 * acc_13 
accuracy_slope =~ 1 * acc_14 
accuracy_slope =~ 2 * acc_15 
accuracy_slope =~ 3 * acc_16 
accuracy_slope =~ 4 * acc_17
accuracy_slope =~ 5 * acc_18 
accuracy_slope =~ 6 * acc_19
accuracy_slope =~ 7 * acc_20
accuracy_slope =~ 8 * acc_21
accuracy_slope =~ 9 * acc_22
accuracy_slope =~ 10 * acc_23
accuracy_slope =~ 11 * acc_24
accuracy_slope =~ 12 * acc_25
accuracy_slope =~ 13 * acc_26

rt_intercept =~ 1 * rt_13 
rt_intercept =~ 1 * rt_14 
rt_intercept =~ 1 * rt_15 
rt_intercept =~ 1 * rt_16 
rt_intercept =~ 1 * rt_17
rt_intercept =~ 1 * rt_18 
rt_intercept =~ 1 * rt_19
rt_intercept =~ 1 * rt_20
rt_intercept =~ 1 * rt_21
rt_intercept =~ 1 * rt_22
rt_intercept =~ 1 * rt_23
rt_intercept =~ 1 * rt_24
rt_intercept =~ 1 * rt_25
rt_intercept =~ 1 * rt_26

rt_slope =~ 0 * rt_13 
rt_slope =~ 1 * rt_14 
rt_slope =~ 2 * rt_15 
rt_slope =~ 3 * rt_16 
rt_slope =~ 4 * rt_17
rt_slope =~ 5 * rt_18 
rt_slope =~ 6 * rt_19
rt_slope =~ 7 * rt_20
rt_slope =~ 8 * rt_21
rt_slope =~ 9 * rt_22
rt_slope =~ 10 * rt_23
rt_slope =~ 11 * rt_24
rt_slope =~ 12 * rt_25
rt_slope =~ 13 * rt_26

# factor variances
accuracy_intercept ~~ accuracy_intercept
accuracy_slope ~~ accuracy_slope
rt_intercept ~~ rt_intercept
rt_slope ~~ rt_slope

# covariances among factors 
accuracy_intercept ~~ rt_intercept
accuracy_slope ~~ rt_slope
accuracy_intercept ~~ accuracy_slope
rt_intercept ~~ rt_slope

#residual covariances
rt_13 ~~ sigma * acc_13
rt_14 ~~ sigma * acc_14 
rt_15 ~~ sigma * acc_15 
rt_16 ~~ sigma * acc_16 
rt_17 ~~ sigma * acc_17
rt_18 ~~ sigma * acc_18 
rt_19 ~~ sigma * acc_19
rt_20 ~~ sigma * acc_20
rt_21 ~~ sigma * acc_21
rt_22 ~~ sigma * acc_22
rt_23 ~~ sigma * acc_23
rt_24 ~~ sigma * acc_24
rt_25 ~~ sigma * acc_25
rt_26 ~~ sigma * acc_26

# means
accuracy_intercept ~ 1
accuracy_slope ~ 1
rt_intercept ~ 1
rt_slope ~ 1

# manifest variances (made equivalent by naming theta)
acc_13 ~~ theta * acc_13 
acc_14 ~~ theta * acc_14 
acc_15 ~~ theta * acc_15 
acc_16 ~~ theta * acc_16 
acc_17 ~~ theta * acc_17
acc_18 ~~ theta * acc_18 
acc_19 ~~ theta * acc_19
acc_20 ~~ theta * acc_20
acc_21 ~~ theta * acc_21
acc_22 ~~ theta * acc_22
acc_23 ~~ theta * acc_23
acc_24 ~~ theta * acc_24
acc_25 ~~ theta * acc_25
acc_26 ~~ theta * acc_26

rt_13 ~~ theta2 * rt_13
rt_14 ~~ theta2 * rt_14 
rt_15 ~~ theta2 * rt_15 
rt_16 ~~ theta2 * rt_16 
rt_17 ~~ theta2 * rt_17
rt_18 ~~ theta2 * rt_18 
rt_19 ~~ theta2 * rt_19
rt_20 ~~ theta2 * rt_20
rt_21 ~~ theta2 * rt_21
rt_22 ~~ theta2 * rt_22
rt_23 ~~ theta2 * rt_23
rt_24 ~~ theta2 * rt_24
rt_25 ~~ theta2 * rt_25
rt_26 ~~ theta2 * rt_26

# manifest means (fixed at zero)
acc_13 ~ 0*1
acc_14 ~ 0*1
acc_15 ~ 0*1
acc_16 ~ 0*1
acc_17 ~ 0*1
acc_18 ~ 0*1
acc_19 ~ 0*1
acc_20 ~ 0*1
acc_21 ~ 0*1
acc_22 ~ 0*1
acc_23 ~ 0*1
acc_24 ~ 0*1
acc_25 ~ 0*1
acc_26 ~ 0*1

rt_13 ~ 0*1 
rt_14 ~ 0*1 
rt_15 ~ 0*1 
rt_16 ~ 0*1 
rt_17 ~ 0*1
rt_18 ~ 0*1 
rt_19 ~ 0*1
rt_20 ~ 0*1
rt_21 ~ 0*1
rt_22 ~ 0*1
rt_23 ~ 0*1
rt_24 ~ 0*1
rt_25 ~ 0*1
rt_26 ~ 0*1"

fit_cc <- lavaan(coupled_change, 
                 data = d_wide,
                 std.lv=TRUE, missing='fiml')

summary(fit_cc)
```

```{r}
layout_cc = matrix(
  nrow=4, 
  ncol = 16,
  data = c("rt_intercept","rt_slope", NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
           NA, NA, "rt_13","rt_14","rt_15","rt_16","rt_17","rt_18","rt_19","rt_20","rt_21","rt_22","rt_23","rt_24","rt_25","rt_26",
           NA, NA, "acc_13","acc_14","acc_15","acc_16","acc_17","acc_18","acc_19","acc_20","acc_21","acc_22","acc_23","acc_24","acc_25","acc_26",
           "accuracy_intercept","accuracy_slope", NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA),
  byrow = TRUE)

graph_sem(model = fit_cc, text_size = 3, layout = layout_cc)
```

## Draft 4: Another version of coupled change

This one doesn't have:

* single residual variance
* coupled covariances
* pinned means

```{r}
coupled_change <- "
accuracy_intercept =~ 1 * acc_13 + 1 * acc_15 + 1 * acc_16 + 1 * acc_17 + 1 * acc_18 + 1 * acc_19 + 1 * acc_20 + 1 * acc_21 + 1 * acc_22 + 1 * acc_23 + 1 * acc_24 + 1 * acc_25 + 1 * acc_26

accuracy_slope =~ 0 * acc_13  + 1 * acc_14  + 2 * acc_15  + 3 * acc_16  + 4 * acc_17 + 5 * acc_18  + 6 * acc_19 + 7 * acc_20 + 8 * acc_21 + 9 * acc_22 + 10 * acc_23 + 11 * acc_24 + 12 * acc_25 + 13 * acc_26

rt_intercept =~ 1 * rt_13 + 1 * rt_14  + 1 * rt_15  + 1 * rt_16  + 1 * rt_17 + 1 * rt_18  + 1 * rt_19+ 1 * rt_20 + 1 * rt_21 + 1 * rt_22 + 1 * rt_23 + 1 * rt_24 + 1 * rt_25 + 1 * rt_26

rt_slope =~ 0 * rt_13 + 1 * rt_14 + 2 * rt_15 + 3 * rt_16 + 4 * rt_17+ 5 * rt_18 + 6 * rt_19 + 7 * rt_20 + 8 * rt_21 + 9 * rt_22 + 10 * rt_23 + 11 * rt_24 + 12 * rt_25 + 13 * rt_26"

fit_cc <- growth(coupled_change, data = d_wide, missing='fiml')

summary(fit_cc)
```

```{r}
graph_sem(model = fit_cc, text_size = 3, layout = layout_cc)
```

## Draft 5: Adding vocab to the mix

I was trying to figure out how to link the observed variables, but I think the observations are too sparse.

```{r eval=FALSE}
cc_vocab <- "
accuracy_intercept =~ 1 * acc_13 + 1 * acc_15 + 1 * acc_16 + 1 * acc_17 + 1 * acc_18 + 1 * acc_19 + 1 * acc_20 + 1 * acc_21 + 1 * acc_22 + 1 * acc_23 + 1 * acc_24 + 1 * acc_25 + 1 * acc_26

accuracy_slope =~ 0 * acc_13  + 1 * acc_14  + 2 * acc_15  + 3 * acc_16  + 4 * acc_17 + 5 * acc_18  + 6 * acc_19 + 7 * acc_20 + 8 * acc_21 + 9 * acc_22 + 10 * acc_23 + 11 * acc_24 + 12 * acc_25 + 13 * acc_26

rt_intercept =~ 1 * rt_13 + 1 * rt_14  + 1 * rt_15  + 1 * rt_16  + 1 * rt_17 + 1 * rt_18  + 1 * rt_19+ 1 * rt_20 + 1 * rt_21 + 1 * rt_22 + 1 * rt_23 + 1 * rt_24 + 1 * rt_25 + 1 * rt_26

rt_slope =~ 0 * rt_13 + 1 * rt_14 + 2 * rt_15 + 3 * rt_16 + 4 * rt_17+ 5 * rt_18 + 6 * rt_19 + 7 * rt_20 + 8 * rt_21 + 9 * rt_22 + 10 * rt_23 + 11 * rt_24 + 12 * rt_25 + 13 * rt_26

prod_intercept =~ 1 * prod_13 + 1 * prod_14  + 1 * prod_15  + 1 * prod_16  + 1 * prod_17 + 1 * prod_18  + 1 * prod_19+ 1 * prod_20 + 1 * prod_21 + 1 * prod_22 + 1 * prod_23 + 1 * prod_24 + 1 * prod_25 + 1 * prod_26

prod_slope =~ 0 * prod_13 + 1 * prod_14 + 2 * prod_15 + 3 * prod_16 + 4 * prod_17+ 5 * prod_18 + 6 * prod_19 + 7 * prod_20 + 8 * prod_21 + 9 * prod_22 + 10 * prod_23 + 11 * prod_24 + 12 * prod_25 + 13 * prod_26"

fit_ccv <- growth(cc_vocab, data = d_wide, missing='fiml', 
                  optim.method = "GN", 
                  optim.gn.tol.x = .001, 
                  optim.gn.iter.max = 500
)

summary(fit_ccv)
```

This model does not converge. Let's think about other ways to describe the same coupling.

## Draft 6: Longitudinal common factor model

Let's think about a two-point factor model, where we:
* analyze all the data
* average observations 0-2 months apart and then analyze 3-6 month growth

First prep the data for this.

```{r}
d_sub_long <- d_sub |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
                         "long_window_acc_var", "prod", "comp")),
                ~ age_scale(.x,age)) |>
  group_by(subject_id) |>
  arrange(administration_id) |>
  mutate(admin_num = 1:n(), 
         delta_t = c(0,diff(age)))

d_sub_long_t <- d_sub_long |> 
  mutate(t = ifelse(delta_t > 2, "t2", "t1")) |>
  group_by(subject_id, dataset_name, t) |>
  summarise(across(all_of(c("rt", "rt_var", "long_window_accuracy",
                            "long_window_acc_var", "prod", "comp")),
                   ~ mean(.x, na.rm=TRUE))) |> 
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy",
                         "long_window_acc_var", "prod", "comp")),
                ~ ifelse(is.nan(.x), NA, .x)))

d_sub_wide <- d_sub_long_t |>
  pivot_wider(id_cols = c("subject_id","dataset_name"), 
              names_from = "t",
              values_from = c("rt", "rt_var", "long_window_accuracy", 
                              "long_window_acc_var", "prod", "comp"))
```

Big choice point here is whether to constrain the loadings, which makes something wrong but interpretable.

```{r}
fa3_model_long <- "
# measurement model
vocab_t1 =~ 1*prod_t1 + 1*comp_t1
accuracy_t1 =~ 1*long_window_accuracy_t1 + 1*long_window_acc_var_t1
speed_t1 =~ 1*rt_t1 + 1*rt_var_t1

vocab_t2 =~ 1*prod_t2 + 1*comp_t2
accuracy_t2 =~ 1*long_window_accuracy_t2 + 1*long_window_acc_var_t2
speed_t2 =~ 1*rt_t2 + 1*rt_var_t2

# longitudinal relationships
vocab_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1

# factor covariances
vocab_t1 ~~ accuracy_t1
vocab_t1 ~~ speed_t1
accuracy_t1 ~~ speed_t1

vocab_t2 ~~ accuracy_t2
vocab_t2 ~~ speed_t2
accuracy_t2 ~~ speed_t2
"

fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')

summary(fit3_long, fit.measures=TRUE, standardize=TRUE)

layout_long = matrix(nrow=5, ncol = 7,
                     data = c("rt_var_t1","rt_t1", "long_window_accuracy_t1", "long_window_acc_var_t1", "prod_t1","comp_t1", NA,
                              "speed_t1",NA,"accuracy_t1",NA,"vocab_t1",NA,  NA,
                              NA, NA, NA, NA, NA, NA,  NA,
                               NA,"speed_t2",NA,"accuracy_t2",NA,"vocab_t2",NA,
                               NA,"rt_var_t2","rt_t2", "long_window_accuracy_t2", "long_window_acc_var_t2", "prod_t2","comp_t2"), 
                     byrow = TRUE)
graph_sem(model = fit3_long, text_size = 3, layout = t(layout_long))

```
To test for fit:

- Constrain regression coefficients to be zero and test for misfit (e.g., key fernald coefficient)
- those are nested so chi2 test can compare between them 
- maybe constrain ALL cross-influences to zero

Alternative with no identification constraints:

```{r}
fa3_model_long <- "
# measurement model
vocab_t1 =~ 1*prod_t1 + comp_t1
accuracy_t1 =~ 1*long_window_accuracy_t1 + long_window_acc_var_t1
speed_t1 =~ 1*rt_t1 + rt_var_t1

vocab_t2 =~ 1*prod_t2 + comp_t2
accuracy_t2 =~ 1*long_window_accuracy_t2 + long_window_acc_var_t2
speed_t2 =~ 1*rt_t2 + rt_var_t2

# longitudinal relationships
vocab_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1

# factor covariances
vocab_t1 ~~ accuracy_t1
vocab_t1 ~~ speed_t1
accuracy_t1 ~~ speed_t1

vocab_t2 ~~ accuracy_t2
vocab_t2 ~~ speed_t2
accuracy_t2 ~~ speed_t2

# means for the latents
vocab_t1 ~ 1
vocab_t2 ~ 1
accuracy_t1 ~ 1
accuracy_t2 ~ 1
speed_t1 ~ 1
speed_t2 ~ 1

# residual variance for the latents
vocab_t1 ~~ NA*vocab_t1
vocab_t2 ~~ NA*vocab_t2
accuracy_t1 ~~ NA*accuracy_t1
accuracy_t2 ~~ NA*accuracy_t2
speed_t1 ~~ NA*speed_t1
speed_t2 ~~ NA*speed_t2
"

fit3_long <- sem(fa3_model_long, d_sub_wide, std.lv=TRUE, missing='fiml')

summary(fit3_long, fit.measures=TRUE, standardize=TRUE)

graph_sem(model = fit3_long, text_size = 3, layout = t(layout_long))
```

Now with baseline compariosns

```{r}
fa3_model_long_baseline <- "
# measurement model
vocab_t1 =~ 1*prod_t1 + comp_t1
accuracy_t1 =~ 1*long_window_accuracy_t1 + long_window_acc_var_t1
speed_t1 =~ 1*rt_t1 + rt_var_t1

vocab_t2 =~ 1*prod_t2 + comp_t2
accuracy_t2 =~ 1*long_window_accuracy_t2 + long_window_acc_var_t2
speed_t2 =~ 1*rt_t2 + rt_var_t2

# longitudinal relationships
vocab_t2 ~ vocab_t1 + 0 * speed_t1 + 0 * accuracy_t1
speed_t2 ~ 0 * vocab_t1 + speed_t1 + 0 * accuracy_t1
accuracy_t2 ~ 0 * vocab_t1 + 0 * speed_t1 + accuracy_t1

# factor covariances
vocab_t1 ~~ accuracy_t1
vocab_t1 ~~ speed_t1
accuracy_t1 ~~ speed_t1

vocab_t2 ~~ accuracy_t2
vocab_t2 ~~ speed_t2
accuracy_t2 ~~ speed_t2

# means for the latents
vocab_t1 ~ 1
vocab_t2 ~ 1
accuracy_t1 ~ 1
accuracy_t2 ~ 1
speed_t1 ~ 1
speed_t2 ~ 1

# residual variance for the latents
vocab_t1 ~~ NA*vocab_t1
vocab_t2 ~~ NA*vocab_t2
accuracy_t1 ~~ NA*accuracy_t1
accuracy_t2 ~~ NA*accuracy_t2
speed_t1 ~~ NA*speed_t1
speed_t2 ~~ NA*speed_t2
"

fit3_long_baseline <- sem(fa3_model_long_baseline, 
                          d_sub_wide, std.lv=TRUE, missing='fiml')

summary(fit3_long_baseline, fit.measures=TRUE, standardize=TRUE)

anova(fit3_long, fit3_long_baseline)
```
With a close baseline.

```{r}
fa3_model_long_closebaseline <- "
# measurement model
vocab_t1 =~ 1*prod_t1 + comp_t1
accuracy_t1 =~ 1*long_window_accuracy_t1 + long_window_acc_var_t1
speed_t1 =~ 1*rt_t1 + rt_var_t1

vocab_t2 =~ 1*prod_t2 + comp_t2
accuracy_t2 =~ 1*long_window_accuracy_t2 + long_window_acc_var_t2
speed_t2 =~ 1*rt_t2 + rt_var_t2

# longitudinal relationships
vocab_t2 ~ vocab_t1 + 0* speed_t1 + accuracy_t1
speed_t2 ~ vocab_t1 + speed_t1 + accuracy_t1
accuracy_t2 ~ vocab_t1 + speed_t1 + accuracy_t1

# factor covariances
vocab_t1 ~~ accuracy_t1
vocab_t1 ~~ speed_t1
accuracy_t1 ~~ speed_t1

vocab_t2 ~~ accuracy_t2
vocab_t2 ~~ speed_t2
accuracy_t2 ~~ speed_t2

# means for the latents
vocab_t1 ~ 1
vocab_t2 ~ 1
accuracy_t1 ~ 1
accuracy_t2 ~ 1
speed_t1 ~ 1
speed_t2 ~ 1

# residual variance for the latents
vocab_t1 ~~ NA*vocab_t1
vocab_t2 ~~ NA*vocab_t2
accuracy_t1 ~~ NA*accuracy_t1
accuracy_t2 ~~ NA*accuracy_t2
speed_t1 ~~ NA*speed_t1
speed_t2 ~~ NA*speed_t2
"
